---
title: "HW2-exploration"
author: "Sam Muir"
date: 2024-03-02
format:
  html:
    embed-resources: true
---

# Part 2a.

```{r, libraries, message=FALSE}
# libraries
library(metajam) 
library(tidyverse)
library(janitor)
```

```{r, read data, message=FALSE}
#...................download data from DataOne...................
# download_d1_data_pkg("https://gmn.lternet.edu/mn/v2/packages/application%2Fbagit-097/doi%3A10.6073%2Fpasta%2F9746c71b35e9f8c544ea12c601c33949", path = here::here("data")) 

#........................read in data.......................
# the data was formatted very weirdly with a few different column headers, so first 2 rows needed to be not read in (skip = 2), the third row was the correct column header, and then there were 2 more rows after that which were other info on the data types, so they were removed as well [-c(1,2),]

animals <- read_csv("data/animals/PLT-GCEM-1610_Animals_9_0.CSV", skip = 2)[-c(1,2),]
biomass <- read_csv("data/biomass/PLT-GCEM-1610_Biomass_10_0.CSV", skip = 2)[-c(1,2),]
vegetation <- read_csv("data/vegetation/PLT-GCEM-1610_Vegetation_9_0.CSV", skip = 2)[-c(1,2),]
stemheight <- read_csv("data/stemheight/PLT-GCEM-1610_StemHeight_10_0.CSV", skip = 2)[-c(1,2),]
```

```{r, data cleaning, warning=FALSE, message=FALSE}
# data cleaning 
# but now after all of the wrangling to get the data read in right, all of the columns are classified as characters...

animals_clean <- animals %>%
  clean_names() %>%
  mutate_at(c("latitude", "longitude", "littoraria_count"), as.numeric) %>% # change cols to numeric
  mutate_at(c("grasshopper_abundance_index", "prokelisia_abundance_index", "littoraria_quadrat_area", "plot_number", "spartina_zone"), as.factor) %>% # change cols to factor
  mutate(date = lubridate::as_date(date)) # change date to a date

biomass_clean <- biomass %>%
  clean_names() %>%
  mutate_at(c("biomass_aboveground_green", "biomass_aboveground_brown", "biomass_aboveground_total", "biomass_belowground_0to10cm", "biomass_belowground_10to30cm", "biomass_belowground_total", "rhizomes_0to10cm", "root_to_shoot_ratio"), as.numeric) %>% # change cols to numeric
  mutate_at(c("spartina_zone", "replicate"), as.factor) %>% # change cols to factor
  mutate(date = lubridate::as_date(date)) # change date to a date

vegetation_clean <- vegetation %>%
  clean_names() %>%
  mutate_at(c("latitude", "longitude", "plant_height"), as.numeric) %>% # change cols to numeric
  mutate_at(c("spartina_zone", "plot_number", "quadrat_area", "flowering_status"), as.factor) %>% # change cols to factor
  mutate(date = lubridate::as_date(date), # change date to a date
         year = as.factor(year(date))) # get the year from the date, change to a factor

stemheight_clean <- stemheight %>%
  clean_names() %>%
  mutate_at(c("spartina_zone"), as.factor) %>% # change cols to factor
  mutate_at(c("core_number", "stem_number", "stem_height"), as.numeric) %>% # change cols to numeric
  mutate(date = lubridate::as_date(date))

## maybe there was a better way to do all of this? but seems like good practice either way 

# joining some of the data frames
stem_animals <- full_join(animals_clean, stemheight_clean, relationship = "many-to-many") %>% # join animals and stemheight
  mutate(year = as.factor(year(date)), # get year from the date
         month = as.factor(month(date))) %>% # get month from the date
  full_join(biomass_clean)
```

## exploratory data viz

**initial question:**    
- how do littoraria counts change over time and how do they vary over different spartina zones?   
- how does biomass vary over time?   

```{r, warning=FALSE, message=FALSE}
# littoratia counts
ggplot(animals_clean, aes(x = littoraria_count)) +
  geom_histogram()

# stem height
ggplot(stemheight_clean, aes(x = stem_height)) +
  geom_histogram()

# stem height + littoraria counts
ggplot(stem_animals, aes(x = stem_height, y = littoraria_count)) +
  geom_point() +
  facet_wrap(~spartina_zone)

# counts by flowering status
vegetation_clean %>%
  group_by(spartina_zone, flowering_status, year) %>% # define groupings
  summarise(count = n()) %>% # count the number of rows
  ggplot(., aes(x = year, y = count, fill = flowering_status)) +
  geom_col(position = "dodge") + # bars next to each other 
  facet_wrap(~spartina_zone)
```

### Plots I'm interested in improving on

```{r, warning=FALSE, message=FALSE}
# littoraria count by spartina zone over time
ggplot(stem_animals, aes(x = year, y = littoraria_count, fill = spartina_zone)) +
  geom_col(position = 'dodge') + # bars next to each other
  facet_wrap(~spartina_zone) +
  theme_linedraw() +
  scale_fill_manual(values = c("seagreen", "cornflowerblue", "magenta4"))

# same as above but line plot
ggplot(stem_animals, aes(x = date, y = littoraria_count, color = spartina_zone)) +
  geom_line() +
  facet_wrap(~spartina_zone, ncol = 1) +
  theme_linedraw() +
  scale_color_manual(values = c("seagreen", "cornflowerblue", "magenta4"))

# biomass by spartina zone over time
ggplot(stem_animals, aes(x = spartina_zone, y = biomass_aboveground_total, fill = year)) +
  geom_boxplot() +
  scale_color_manual(values = c("seagreen", "cornflowerblue", "magenta4"))
```

## Part 2b.

1.
The thing I found interesting (but now thinking about it makes sense) is that there is a lot of 0s for the *littoraria* counts. I did also not expect the other animals counts to be binary (0 not present, 1 present) and I'm not sure the best way to visualize binary data like this. From looking at some of my exploratory visualizations, I can see there are differences in *littoraria* counts across different *spartina* zones (short, medium, and tall). For short and tall there also seems to be a general decrease in *littoraria* counts over time, but for medium zones, there is not a clear pattern. I also learned that there were a lot more instances of plants not flowering rather than flowering. From exploring the data, I think I am more interested in exploring the *littoraria* counts.

2. 
Yes and no. I think that now seeing how the data types, I am wanting to shift my question(s) to how *littoraria* counts vary by *spartina* zone overtime and above ground biomass over time. I think it would be difficult to mix continuous counts and binary occurrences. For the line plots over time that I have made above, I am thinking that instead of plotting date or just year on the x-axis, it might be better to plot month or month-year. It looks like that data was collected in each month over the survey years, but not on a consistent date. I would think that there would be some kind of seasonal effect, so it could be more interesting to see how these counts/ measurements vary month to month rather than year to year. With adjusting my questions a bit, I am happy with the initial visualizations here and I don't think that I need any more data. I still think that looking at the data spatially would be interesting as well since there are coordinates for each sample, but this might not be necessary for the questions I'm interested in.

3. 
One of the big challenges with the data was getting it read in correctly. It had a lot of different rows before the actual data began that was information about the collection and data types. This made it hard to read in correctly. Another challenge will be deciding what variables are most important to compare to each other and how to present that (ie. a visualization that makes it easier to compare counts by zone or a viz that makes it easier to compare counts over time). If I want to summarize counts by month for each year, I will need to take some time to get everything wrangled and clean for that. Plotting the points by the coordinates will also take some data wrangling.  

